How to publish in Biological Sciences: a guide for the uninitiated

### 2022-09-08 1.09-preprints.Rmd
# update on JMRI

Surely, the onset of 'Overlay Journals' should have publishers quaking in their boots? Strangley not. But their response should really be enough to wake us up.

> "They're probably only going to succeed in disciplines where a no-frills approach to publishing is acceptable... 
>
> I think that the real threat to our traditional model... is if Overlay Journals have **Impact Factors** and can provide the same services, and they are free... then I think that that does pose a threat."
>
> ---[Anonymous from Highwire Press Inc. 2017](https://youtu.be/ADecAXLrKoA)

@@

### What is happening in biological sciences? {#PCI-EvolBiol}

One of the original electronic journals, [*Journal of Medical Internet Research* (*JMIR*)](https://www.jmir.org/), announced in 2019 that it will launch an overlay journal covering biology, a so called 'superjournal', *JMIRx* [@eysenbach2019celebrating]. This overlay journal operates by editors choosing preprints that they want to publish ('editorial prospecting'), and then approaching authors and reviewers, and also by authors pitching their preprints to the editors. Today [*JMIRx|Bio*](https://bio.jmirx.org/) accepts any preprint published on [*bioRxiv*](#bioRxiv). Although *JMIRx|Bio*, and sister journal [*JMRIx|Psy*](https://psy.jmirx.org/), were launched in 2020, I cannot find any articles submitted (by mid-2021). The sister journal [*JMIRx|Med*](https://med.jmirx.org/), launched at the same time and in the same area as other JMIR journals, and has a rapidly expanding publication base.

@@

All of these models tie-in with [transparency](#transparency). Because preprints and reviews are OA, the review process is transparent.

#### Peer Community In

@@

A nearly 'overlay' model is [*Peer Community in Evolutionary Biology*](https://evolbiol.peercommunityin.org/), launched in January 2017. This comes very close to the '*arXiv* Overlay Journal' model described above. These preprints are submitted to PCI-Evol Biol, and are reviewed and (if they aren't rejected), a recommendation is given. The site then publishes the recommendation from peers as well as pointing to the preprint. However, unlike _Discrete Analysis_, the preprint remains 'unpublished' despite the peer review and can then be taken onto a traditional journal. There is a  growing [list of journals](https://evolbiol.peercommunityin.org/about/about) whose editors will accept recommendations from PCI-EvolBiol, and may use the reviews when appropriate. However, it's also worth noting that there are a small number of journals that will not accept preprints recommended by PCI-EvolBiol. While *Peer Community in Evolutionary Biology* does not publish their peer reviewed articles, another initiative from *Peer Community In* takes a step backwards to get a step closer ([see below](#commitment)).

@@

### Beware of publishers' preprint servers {#beware}
Publishers have a lot to be worried about preprint servers as they have the potential to take away their business. For this reason, you will see that some publishers have launched their own preprint services. Don't be lured into submitting to the preprint server of a publisher. There are plenty of preprint servers that have nothing to do with publishers, including the most ubiquitous. These are non-profit transparent organisations, and we all have an interest in them staying that way. Simply put, there's no need to use a publishers' preprint server or other tools that they offer. Instead use Open Source Tools and transparent resources [@kramer2016innovations] to avoid [academic capture](#capture).

>>>>

Surely, the onset of 'Overlay Journals' should have publishers quaking in their boots? Strangely not. But their response should really be enough to wake us up.

> "They're probably only going to succeed in disciplines where a no-frills approach to publishing is acceptable... 
>
> I think that the real threat to our traditional model... is if Overlay Journals have **Impact Factors** and can provide the same services, and they are free... then I think that that does pose a threat."
>
> ---[Claire Rawlinson, Highwire Press Inc. 2017](https://youtu.be/ADecAXLrKoA)

@@

### What is happening in biological sciences?

Unfortunately, the good qualities of overlay journals are being exploited by some publishers without any follow through that produces [Diamond OA](#diamond) content.

One of the original electronic journals, [*Journal of Medical Internet Research* (*JMIR*)](https://www.jmir.org/), announced in 2019 that it will launch an overlay journal covering biology, a so called 'superjournal', *JMIRx* [@eysenbach2019celebrating]. However, JMIR appears to be exploiting the name 'overlay journal' without adhering to any of the qualities - they are essentially more [Gold OA](#gold) journals with ['no impact factor'](#noif) aspirations that intend to use the [Article Processing Charge](#APC) to finance their operation. Editors may use preprints to approach authors of papers that they want to publish ('editorial prospecting'), and then approaching reviewers, and also by authors pitching their preprints to the editors. Today [*JMIRx|Bio*](https://bio.jmirx.org/) accepts any preprint published on [*bioRxiv*](#bioRxiv). Although *JMIRx|Bio*, and sister journal [*JMRIx|Psy*](https://psy.jmirx.org/), were launched in 2020, I cannot find any articles submitted (by mid-2022). The sister journal [*JMIRx|Med*](https://med.jmirx.org/), launched at the same time and in the same area as other JMIR journals, and has a rapidly expanding publication base.  But if you want to publish in JMIR you can expect to pay an [APC](#APC), and so they are not overlay journals as they archive their own content. But JMIR is disrupting the publishing scene and offers a lot of benefits towards open science.

@@

All of these models tie-in with [transparency](#transparency). Because preprints and reviews are OA, and the review process is transparent.

#### Peer Community In {#PCI-EvolBiol}

@@

A nearly 'overlay' model is [*Peer Community in Evolutionary Biology*](https://evolbiol.peercommunityin.org/), launched in January 2017. This comes very close to the '*arXiv* Overlay Journal' model described above. These preprints are submitted to *PCI-Evol Biol*, and are reviewed and (if they aren't rejected), a recommendation is given. The site then publishes the recommendation from peers as well as pointing to the preprint. However, unlike _Discrete Analysis_, the preprint remains 'unpublished' despite the peer review and can then be taken onto a traditional journal. Since the success of *PCI-Evol Biol*, the list of Peer Communities has grown to include *Peer Communities In: Evolutionary Biology, Ecology, Paleontology, Animal Science, Zoology, Neuroscience, Genomics and Mathematical & Computational Biology.* Find a complete list of current Peer Communities In [here](https://peercommunityin.org/current-pcis/).

There is a  growing [list of journals](https://evolbiol.peercommunityin.org/about/about) whose editors will accept recommendations from PCIs, and may use the reviews when appropriate. However, it's also worth noting that there are a small number of journals that will not accept preprints recommended by PCIs.

@@

### Choose non-profit preprint servers {#beware}
Preprint servers have the potential to take away traditional publishers' business. For this reason, you will see that some publishers have launched their own preprint services. However, there are plenty of preprint servers that have nothing to do with publishers, including the most ubiquitous. These are non-profit transparent organisations, and we all have an interest in them staying that way. Instead use Open Source Tools and transparent resources [@kramer2016innovations].
